---
title: "Getting started with onepagr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with onepagr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
quarto_ready <- requireNamespace("quarto", quietly = TRUE) &&
  quarto::quarto_available()
pdftools_ready <- quarto_ready && requireNamespace("pdftools", quietly = TRUE)
```

onepagr turns a named list of values into a finished, accessible PDF
report using one of its built-in Typst templates. This vignette walks
through rendering your first report, picking a theme, and moving on to
customizing a template.

## Before you start

onepagr needs [Quarto](https://quarto.org) (which bundles Typst) on your
system. Check whether it's available:

```{r}
onepagr::check_quarto()
```

If that reports a problem, install a user-local copy yourself with
`onepagr::install_quarto()` (onepagr never installs anything
automatically) or install Quarto normally from quarto.org, then restart
your R session.

On macOS and Linux, `install_quarto()` also points onepagr at the copy
it just installed: it sets `QUARTO_PATH` for your current session right
away, then asks whether to save that to your `~/.Renviron` too, so every
future session picks it up automatically. You're never expected to
hand-edit a `.Renviron` file yourself. (Windows opens the official
installer instead, so there's no path to set until you finish that
yourself.)

If you already have a Quarto install elsewhere onepagr should use
instead (an admin-managed one on Posit Workbench, for example),
`onepagr::set_quarto_path("/path/to/quarto")` points onepagr at it the
same way, with the same save-for-later prompt.

## Render your first report

Every template needs a named list of values, and which values depend on
the template. `template_data()` gives you a complete, working starting
point for any built-in template: every value is already filled in, a
real one, not a blank you have to guess the shape of.

```{r, eval = quarto_ready}
library(onepagr)

data <- template_data("cohort_summary")
out <- tempfile(fileext = ".pdf")
render_onepager(
  data, template = "cohort_summary", theme = "default", output = out
)
file.exists(out)
```

That's the whole workflow: a starter list, a finished PDF. `data` here
is the package's own example data (see `example_data()`), so this
renders exactly the PDF shown below in "What you get" without changing
a thing. A real project replaces the values it has real numbers for:

```{r, eval = FALSE}
data$n_decedents <- "6,000"
data$strip_period <- "2025"
# Every heading, label and paragraph keeps its default wording unless
# you set it too -- see vignette("theming"), Part 5.
```

`attr(data, "required")` names the entries `template_data()` had no
default for -- the ones a real project actually needs its own numbers
for, as opposed to wording you can leave alone:

```{r}
attr(data, "required")
```

You'd typically build those values from your analysis output using
ordinary R (`sprintf()`, `scales::comma()`, `paste()`), the same way
you'd assemble any other report's numbers -- not hand-type them the way
this vignette's example does.

## What you get

This is the actual PDF the `render_onepager()` call above produces
(front page; `cohort_summary` is a fixed two-page template, so a
matching back page follows it in the real file):

```{r, eval = pdftools_ready, echo = FALSE}
# Written under knitr's figure path, not tempdir(): pkgdown copies figures
# after the article renders, by which point a tempdir() file is gone.
front_page <- knitr::fig_path(".png")
dir.create(dirname(front_page), recursive = TRUE, showWarnings = FALSE)
pdftools::pdf_convert(
  out, pages = 1, dpi = 100, filenames = front_page, verbose = FALSE
)
knitr::include_graphics(front_page)
```

By default, `render_onepager()` leaves the resolved `.typ` source
(and everything it needs to recompile: the theme, shared components,
and assets) next to your output PDF, in a folder named after it, here
something like `<tempfile>_typst/`. That's deliberate: the actual Typst
source is never hidden, even if you never open it. If you want only the
PDF and nothing else, pass `keep_typst = FALSE`.

## Picking a theme

onepagr ships three built-in themes, selectable by name:

```{r, eval = FALSE}
render_onepager(data, template = "cohort_summary", theme = "default", output = "report.pdf")
render_onepager(data, template = "cohort_summary", theme = "uk", output = "report.pdf")
render_onepager(data, template = "cohort_summary", theme = "kdph", output = "report.pdf")
```

List what's available:

```{r}
onepagr::list_themes()
```

Your own project can supply a fully custom theme instead. See
`?resolve_theme` for how `theme_path` works, and any file under
`system.file("typst/themes", package = "onepagr")` for the token schema
a custom theme needs to define.

## Exploring or customizing a template

To see a template's real Typst source, or hand-edit one, export it into
your own project rather than reading it out of the installed package:

```{r, eval = FALSE}
export_template("cohort_summary", "my-report/", theme = "uk")
```

That copies the template, the resolved theme, shared components, and
assets into `my-report/` as one self-contained, independently
compilable unit. Once exported, it's yours: onepagr never touches it
again. Compile your edited copy directly with `compile_typst()`.

## What's required for each template

Every template validates its input before compiling: if a required
value is missing, `render_onepager()` raises a clear R error listing
exactly which ones, rather than silently producing a PDF with a blank
spot. You can check what a template needs ahead of time:

```{r}
onepagr::list_templates()
```

```{r}
# Every token a template takes: required ones first, then optional ones
# with their default text.
head(onepagr::template_tokens("overdose_spike_alert"), 8)
```

The optional tokens are the template's own wording (headings, captions,
paragraphs, labels). Set any of them in `data` to reword it; see
`vignette("theming")`, Part 5.

```{r, eval = FALSE}
# Path to a built-in template's source, if you want to read its {{{token}}} usage directly
onepagr::resolve_template("overdose_spike_alert")
```

## Learn more

- `?render_onepager`, `?export_template`, `?compile_typst` for the full
  function reference.
- `?check_quarto`, `?install_quarto` for the Quarto/Typst toolchain
  helpers.
- The package README for the full list of built-in templates and their
  informational shapes.
