---
title: "Getting started with onepagr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with onepagr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
quarto_ready <- requireNamespace("quarto", quietly = TRUE) &&
  quarto::quarto_available()
```

onepagr turns a named list of values into a finished, accessible PDF
report using one of its built-in Typst templates. This vignette walks
through rendering your first report, picking a theme, and moving on to
customizing a template.

## Before you start

onepagr needs [Quarto](https://quarto.org) (which bundles Typst) on your
system. Check whether it's available:

```{r}
onepagr::check_quarto()
```

If that reports a problem, install a user-local copy yourself with
`onepagr::install_quarto()` (onepagr never installs anything
automatically) or install Quarto normally from quarto.org, then restart
your R session.

## Render your first report

Every template needs a named list of values — which values depend on
the template. `cohort_summary` (contrasting two groups at a point in
time) needs the fewest, so it's a good first example.

```{r, eval = quarto_ready}
library(onepagr)

data <- list(
  doc_title = "SAMPLE COHORT SUMMARY",
  doc_subtitle = "Sample Program . Sample Region",
  org_full = "Sample Health Organization",
  contact_url = "https://example.org/",
  contact_email = "contact@example.org",
  strip_data = "Sample System",
  strip_period = "2024",
  strip_design = "Retrospective Cohort",
  strip_geography = "Sample Region",
  n_decedents = "5,267", n_ems_total = "24,906", pct_linked = "84%",
  n_prior_ems = "21,021", n_eligible_decedents = "6,260",
  n_unlinked_decedents = "993", n_coroner_counties = "120",
  pct_linked_male_width = "64.6", n_male = "3,400",
  pct_linked_female_width = "35.4", n_female = "1,867",
  pct_linked_appalachian_width = "31.2", n_appalachian = "1,643",
  pct_linked_nonappalachian_width = "68.8", n_nonappalachian = "3,624",
  pct_linked_age_lt25_width = "8", n_age_lt25 = "421",
  pct_linked_age_25_34_width = "22", n_age_25_34 = "1,159",
  pct_linked_age_35_44_width = "27", n_age_35_44 = "1,422",
  pct_linked_age_45_54_width = "24", n_age_45_54 = "1,264",
  pct_linked_age_55_64_width = "14", n_age_55_64 = "737",
  pct_linked_age_65plus_width = "5", n_age_65plus = "264",
  pct_linked_white_width = "86", n_white = "4,527",
  pct_linked_black_width = "10.7", n_black = "574",
  pct_linked_other_width = "1.3", n_other = "66",
  wc_linked_median = "188", wc_unlinked_median = "121",
  domain_diff_scene = "6.8", domain_diff_history = "5.8",
  domain_diff_drug = "2.4", domain_diff_medication = "1.2",
  domain_diff_mental = "0.8", n_prior_od_ems = "5,189",
  tl1_yr = "2019", tl1_label = "Pilot phase",
  tl2_yr = "2020-2022", tl2_label = "Infrastructure build-out",
  tl3_yr = "2023-Present", tl3_label = "Operational",
  pct_any_prior_enc = "76.9%", pct_od_prior_enc = "38.1%",
  mean_prior_enc = "4", median_prior_enc = "2",
  n_od_ems_denom = "5,189", pct_naloxone_width = "70.2",
  pct_naloxone = "70.2%", pct_no_naloxone_width = "29.8",
  pct_no_naloxone = "29.8%", n_naloxone_enc = "3,643",
  n_no_naloxone_enc = "1,546", pct_decedent_nax = "30.8%",
  timing_denom = "2,009", median_days = "330", timing_iqr = "53-941",
  mean_days = "575.1", pct_30d_width = "20.8", n_30d = "417",
  pct_90d_width = "29.8", n_90d = "598", pct_365d_width = "52.8",
  n_365d = "1,061", pct_gt365d_width = "47.2", n_gt365d = "948",
  lessons_learned_text = "Sample lessons-learned narrative text.",
  disclaimer_text = "Sample disclaimer text for illustration only.",
  footnote_sources = "Sample Data Source"
)

out <- tempfile(fileext = ".pdf")
render_onepager(
  data, template = "cohort_summary", theme = "default", output = out
)
file.exists(out)
```

That's the whole workflow: a named list in, a finished PDF out. A real
project won't hand-type every value like this vignette does for
demonstration — you'd typically build this list from your analysis
output using ordinary R (`sprintf()`, `scales::comma()`,
`paste()`), the same way you'd assemble any other report's numbers.

## What you just got

By default, `render_onepager()` leaves the resolved `.typ` source
(and everything it needs to recompile: the theme, shared components,
and assets) next to your output PDF, in a folder named after it — here,
something like `<tempfile>_typst/`. That's deliberate: the actual Typst
source is never hidden, even if you never open it. If you want only the
PDF and nothing else, pass `keep_typst = FALSE`.

## Picking a theme

onepagr ships two built-in themes, selectable by name:

```{r, eval = FALSE}
render_onepager(data, template = "cohort_summary", theme = "default", output = "report.pdf")
render_onepager(data, template = "cohort_summary", theme = "uk", output = "report.pdf")
```

List what's available:

```{r}
onepagr::list_themes()
```

Your own project can supply a fully custom theme instead — see
`?resolve_theme` for how `theme_path` works, and any file under
`system.file("typst/themes", package = "onepagr")` for the token schema
a custom theme needs to define.

## Exploring or customizing a template

To see a template's real Typst source, or hand-edit one, export it into
your own project rather than reading it out of the installed package:

```{r, eval = FALSE}
export_template("cohort_summary", "my-report/", theme = "uk")
```

That copies the template, the resolved theme, shared components, and
assets into `my-report/` as one self-contained, independently
compilable unit. Once exported, it's yours — onepagr never touches it
again. Compile your edited copy directly with `compile_typst()`.

## What's required for each template

Every template validates its input before compiling: if a required
value is missing, `render_onepager()` raises a clear R error listing
exactly which ones, rather than silently producing a PDF with a blank
spot. You can check what a template needs ahead of time:

```{r}
onepagr::list_templates()
```

```{r, eval = FALSE}
# Path to a built-in template's source, if you want to read its {{{token}}} usage directly
onepagr::resolve_template("overdose_spike_alert")
```

## Learn more

- `?render_onepager`, `?export_template`, `?compile_typst` for the full
  function reference.
- `?check_quarto`, `?install_quarto` for the Quarto/Typst toolchain
  helpers.
- The package README for the full list of built-in templates and their
  informational shapes.
