---
title: "Theming: colors and logos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Theming: colors and logos}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
quarto_ready <- requireNamespace("quarto", quietly = TRUE) &&
  quarto::quarto_available()
```

onepagr's visual identity splits into two genuinely separate mechanisms,
and it's worth knowing which is which before you customize either one:

- **Colors and spacing** are controlled by a **theme**: a single Typst
  dictionary, selectable by name (`theme = "uk"`) or supplied as your own
  file (`theme_path = "my-theme.typ"`).
- **Logos and the header texture** are **data**, not theme settings --
  every template requires seven logo/image tokens in the `data` list you
  pass to `render_onepager()`, the same way it requires `doc_title` or
  `org_full`. Swapping logos means changing those values and staging your
  own image files, not editing a template's Typst source.
- **Fonts** are a theme setting (`body-font`), but making a *specific*
  font file actually available to Typst is a separate, per-render
  concern, covered in Part 3.

This vignette covers all three.

## Part 1: Colors

### Built-in themes

```{r}
onepagr::list_themes()
```

```{r, eval = FALSE}
render_onepager(data, template = "trend_snapshot", theme = "default", output = "report.pdf")
render_onepager(data, template = "trend_snapshot", theme = "uk", output = "report.pdf")
```

### Writing your own theme

A theme is a plain Typst file defining a `theme` dictionary (colors,
fonts, spacing/stroke scale) and a `theme-grad` dictionary (a handful of
gradients derived from `theme`'s colors). The easiest starting point is
a copy of the built-in `default` theme:

```{r}
default_theme_path <- system.file(
  "typst", "themes", "default.typ",
  package = "onepagr"
)
cat(readLines(default_theme_path, n = 6), sep = "\n")
```

Copy that file, then change the values you want. The keys that matter
most for an overall look:

| Key | Controls |
|---|---|
| `brand-blue`, `brand-midnight` | The header band's gradient, and the accent stripe on every stat-card/callout box |
| `brand-accent` | Large decorative text (big stat-card numbers) -- needs 3:1 contrast, not 4.5:1 |
| `brand-accent-text` | The same accent color's small-text-safe variant (e.g. a `text-box` section heading) -- needs 4.5:1 |
| `card-bg`, `callout-bg`, `lessons-bg` | Box background tints |
| `text-secondary`, `text-muted` | Body and caption text colors |
| `severity-warning`, `severity-critical` (+ their `-bg`/`-text` variants) | Alert-template severity coloring (`overdose_spike_alert`, `syndromic_alert`) |
| `map-border-color` | Map-box border stroke (`county_choropleth`) -- kept separate from `brand-midnight` since a real brand spec for this can diverge from it |

The full key list (spacing, stroke widths, radius, margin) is the same
across every theme file; see `default.typ`'s own comments for what each
one does; they're less likely to need changing than the colors above.

### Verifying contrast before you ship a color change

This is the part that's easy to skip and genuinely shouldn't be. Every
color in onepagr's built-in themes was verified against WCAG's real
contrast formula, not chosen by eye, because a color that looks readable
to you is not the same thing as a color that measures 4.5:1. onepagr
doesn't yet ship an automated checker for this (`check_theme_contrast()`
is a planned, not-yet-built function), so for now this is a manual step,
using the same formula this package's own development used:

```{r}
wcag_contrast <- function(hex1, hex2) {
  linearize <- function(channel) {
    channel <- channel / 255
    ifelse(channel <= 0.03928, channel / 12.92, ((channel + 0.055) / 1.055)^2.4)
  }
  relative_luminance <- function(hex) {
    rgb <- grDevices::col2rgb(hex)
    lin <- linearize(rgb)
    0.2126 * lin[1] + 0.7152 * lin[2] + 0.0722 * lin[3]
  }
  l1 <- relative_luminance(hex1)
  l2 <- relative_luminance(hex2)
  (max(l1, l2) + 0.05) / (min(l1, l2) + 0.05)
}

# Example: is a purple accent readable as small text on a light card
# background? WCAG needs >= 4.5 for normal text, >= 3 for large text
# (18pt+ regular or 14pt+ bold).
wcag_contrast("#6A4C93", "#EDE7F6")
```

Check every foreground/background pairing your new color actually
appears in, at the text size it's actually used at (a `text-box` section
label is small text and needs 4.5:1; a 28pt stat-card number is large
text and only needs 3:1). This project's own `default.typ` and `uk.typ`
theme files have real, first-hand examples of a color that passed at one
size and failed at another; both files' comments explain exactly which
pairing failed and why.

### Rendering with a custom theme

```{r, eval = FALSE}
render_onepager(data, template = "trend_snapshot", theme_path = "my-theme.typ", output = "report.pdf")
```

## Part 2: Logos and images

### Logos are data, not template edits

Every built-in template requires these tokens in the `data` list you
pass to `render_onepager()`:

| Token | Meaning |
|---|---|
| `logo_primary_path`, `logo_primary_alt` | Your own organization's logo -- always shown |
| `logo_partner_a_path`, `logo_partner_a_alt`, `show_partner_a` | An optional co-branding partner, shown left of the primary logo |
| `logo_partner_b_path`, `logo_partner_b_alt`, `show_partner_b` | A second optional co-branding partner, shown right of the primary logo |
| `header_texture_path` | The decorative background texture behind the header band |

Changing a logo means changing these values plus staging the actual
image file, via `render_onepager()`'s `extra_assets` argument -- it does
**not** require exporting or hand-editing a template's Typst source.
Three separate logo slots exist (rather than one flattened composite
image) so a screen reader announces real per-organization identification
instead of one opaque "logo" -- but not every jurisdiction has a
three-organization design. `show_partner_a`/`show_partner_b` are each
independent `"true"`/`"false"` toggles (the literal lowercase string, not
an R logical -- see `?render_onepager`), so a single health department
reporting under its own name gets one logo with no dangling divider, a
two-agency partnership gets two, and a KIPRC-style three-organization
lockup gets three. The primary logo is never optional.

```{r, eval = FALSE}
# A single health department: one logo, no partners.
data$logo_primary_path <- "county-health-dept-logo.png"
data$logo_primary_alt <- "Example County Health Department logo"
data$show_partner_a <- "false"
data$show_partner_b <- "false"
data$header_texture_path <- "header-texture.png"

render_onepager(
  data, template = "syndromic_alert", theme = "default", output = "report.pdf",
  extra_assets = c("county-health-dept-logo.png", "header-texture.png")
)
```

`extra_assets` stages your files into the same directory Typst compiles
from (Typst's compiler sandboxes file access and rejects absolute paths
outright), so the token value must be just the basename you pass to
`extra_assets`, not the file's original full path. If you're fine with
the package's own three built-in placeholder logos, you don't need
`extra_assets` at all -- the fixture defaults in
`vignette("getting-started")` already point at those bundled assets by
name.

### A practical tip: crop before you resize

A raster logo file often has transparent padding baked in around the
real artwork (this project found one source logo with roughly a third
of its canvas being empty space). That padding silently shrinks how big
the logo actually looks once Typst applies a `height:`/`width:` setting,
and can misalign a multi-logo row that assumes each image's visible
content fills its frame. Check the image's real bounding box before
trusting a nominal size setting:

```{r, eval = FALSE}
# Using the PIL/Pillow Python library's alpha.getbbox(), or an
# equivalent in your tool of choice, before sizing a new logo.
```

### Accessible alt text

Every logo `image()` call takes an `alt:` argument. Write real,
specific text ("Sample Health Department logo"), not something generic
like "logo" or "image" -- this is what a screen reader announces in
place of the image, and it's the same accessibility bar the rest of
onepagr's output is held to.

## Part 3: Fonts

A theme's `body-font` key names the font family Typst should use for all
body/heading text (e.g. `default.typ`'s `body-font: "Liberation Sans"`).
Two different questions hide inside "change the font":

1. **What family name should Typst look for?** That's the theme's job --
   change `body-font` in your own theme file, same as any other token in
   Part 1.
2. **Is that family actually available to Typst at compile time?** That's
   a separate, per-render concern, and it's the one that actually needs
   new machinery: Typst only sees fonts installed system-wide by default,
   and a locked-down enterprise deployment (e.g. Posit Workbench) may not
   permit installing new fonts at the OS level at all.

`render_onepager()`'s `font_dir` argument solves the second problem
without needing OS-level font installation: point it at a directory
containing your own `.ttf`/`.otf` files, and Typst searches that
directory *in addition to* system fonts for this compile only
(confirmed directly: `quarto typst fonts --font-path <dir>` lists a font
from `<dir>` alongside every system font, not instead of them).

```{r, eval = FALSE}
# my-theme.typ sets body-font: "Open Sans", and Open Sans isn't
# installed system-wide -- ship the font file alongside your project
# and point font_dir at its folder.
render_onepager(
  data, template = "cohort_summary", theme_path = "my-theme.typ",
  output = "report.pdf", font_dir = "fonts/"
)
```

If `body-font` names a family Typst can't find (system-wide or in
`font_dir`), Typst silently falls back to a default font rather than
erroring -- so after changing a font, actually open the compiled PDF (or
check its embedded font list) to confirm your font took effect, the same
"verify, don't assume" discipline Part 1 applies to color contrast.

## Onboarding a new jurisdiction: what you actually need

Pulling Parts 1-3 together, a health department or university adopting
onepagr for the first time needs exactly three things before it can
render its own reports, on top of the same real analysis output every
other onepagr user needs:

1. **A color palette.** At minimum, a primary/accent color pair; see
   Part 1. Reuse the built-in `default` (Bootstrap-derived, brand-neutral)
   theme as-is if you don't have -- or don't yet want to commit to --
   your own institutional colors. A full custom theme is colors, not
   code: copy `default.typ`, change the hex values, verify contrast, done.
2. **Logo file(s).** One, if you're reporting under your own name alone;
   two or three if you co-brand with partner agencies. See Part 2 --
   there's no minimum of three, and no template editing required either
   way.
3. **A font, if the default doesn't suit you.** Optional -- every
   built-in theme already names a real, commonly-available font. Only
   relevant if your institution has a specific brand font it wants to
   use instead; see Part 3.

None of the three requires touching a template's Typst source. A new
jurisdiction's entire setup is a theme file (or none, if `default`/`uk`
suffice), a handful of logo files, and the `data` list documented in
`vignette("getting-started")` and `vignette("end-to-end-workflow")`.

## See also

- `vignette("getting-started", package = "onepagr")` for the basic
  render workflow.
- `vignette("end-to-end-workflow", package = "onepagr")` for a full
  worked example starting from real data.
- Any file under `system.file("typst/themes", package = "onepagr")`
  for a complete, real theme file to read or copy.
