---
title: "Theming: colors and logos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Theming: colors and logos}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
quarto_ready <- requireNamespace("quarto", quietly = TRUE) &&
  quarto::quarto_available()
```

onepagr's visual identity splits into two genuinely separate mechanisms,
and it's worth knowing which is which before you customize either one:

- **Colors and spacing** are controlled by a **theme**: a single Typst
  dictionary, selectable by name (`theme = "uk"`) or supplied as your own
  file (`theme_path = "my-theme.typ"`).
- **Logos and the header texture** are **data**, not theme settings:
  every template requires seven logo/image tokens in the `data` list you
  pass to `render_onepager()`, the same way it requires `doc_title` or
  `org_full`. Swapping logos means changing those values and staging your
  own image files, not editing a template's Typst source.
- **Fonts** are a theme setting (`body-font`), but making a *specific*
  font file actually available to Typst is a separate, per-render
  concern, covered in Part 3.

This vignette covers all three.

## Part 1: Colors

### Built-in themes

```{r}
onepagr::list_themes()
```

```{r, eval = FALSE}
render_onepager(data, template = "trend_snapshot", theme = "default", output = "report.pdf")
render_onepager(data, template = "trend_snapshot", theme = "uk", output = "report.pdf")
render_onepager(data, template = "trend_snapshot", theme = "kdph", output = "report.pdf")
```

Same template, same data, only `theme` changed between the three calls
above (`default`, `uk`, `kdph`, left to right):

```{r, echo = FALSE, fig.show = "hold", out.width = "32%", fig.alt = c("trend_snapshot rendered with the default, brand-neutral theme.", "The same trend_snapshot report rendered with the uk theme instead, showing only colors and typography changed.", "The same trend_snapshot report rendered with the kdph theme: navy header, indigo bars, and Calibri type.")}
knitr::include_graphics(file.path(
  system.file("help", "figures", package = "onepagr"),
  c(
    "example-trend-snapshot-theme-default.png",
    "example-trend-snapshot-theme-uk.png",
    "example-trend-snapshot-theme-kdph.png"
  )
))
```

The `kdph` theme is an unofficial implementation of the Kentucky
Department for Public Health's 2026 Data Visualization Style Guidelines.
Read the header comment in `kdph.typ` before using it for anything
external: the guide's palette is for data visualization and does not
replace KDPH's branding guidelines, logos are supplied through the
`logo_*` tokens (the guide puts the KDPH logo first), and its contrast
numbers were computed and spot-checked with PAC on the sample templates,
so run PAC on your own content before relying on it.

### Writing your own theme

A theme is a plain Typst file defining a `theme` dictionary (colors,
fonts, spacing/stroke scale) and a `theme-grad` dictionary (a handful of
gradients derived from `theme`'s colors). The easiest starting point is
a copy of the built-in `default` theme:

```{r}
default_theme_path <- system.file(
  "typst", "themes", "default.typ",
  package = "onepagr"
)
cat(readLines(default_theme_path, n = 6), sep = "\n")
```

Copy that file, then change the values you want. The keys that matter
most for an overall look:

| Key | Controls |
|---|---|
| `brand-blue`, `brand-midnight` | The header band's gradient, and the accent stripe on every stat-card/callout box |
| `brand-accent` | Large decorative text (big stat-card numbers); needs 3:1 contrast, not 4.5:1 |
| `brand-accent-text` | The same accent color's small-text-safe variant (e.g. a `text-box` section heading); needs 4.5:1 |
| `card-bg`, `callout-bg`, `lessons-bg` | Box background tints |
| `text-secondary`, `text-muted` | Body and caption text colors |
| `severity-warning`, `severity-critical` (+ their `-bg`/`-text` variants) | Alert-template severity coloring (`overdose_spike_alert`, `syndromic_alert`) |
| `map-border-color` | Map-box border stroke (`county_choropleth`); kept separate from `brand-midnight` since a real brand spec for this can diverge from it |

The full key list (spacing, stroke widths, radius, margin) is the same
across every theme file; see `default.typ`'s own comments for what each
one does; they're less likely to need changing than the colors above.

### Verifying contrast before you ship a color change

This is the part that's easy to skip and genuinely shouldn't be. Every
color in onepagr's built-in themes was verified against WCAG's real
contrast formula, not chosen by eye, because a color that looks readable
to you is not the same thing as a color that measures 4.5:1. onepagr
doesn't yet ship an automated checker for this (`check_theme_contrast()`
is a planned, not-yet-built function), so for now this is a manual step,
using the same formula this package's own development used:

```{r}
wcag_contrast <- function(hex1, hex2) {
  linearize <- function(channel) {
    channel <- channel / 255
    ifelse(channel <= 0.03928, channel / 12.92, ((channel + 0.055) / 1.055)^2.4)
  }
  relative_luminance <- function(hex) {
    rgb <- grDevices::col2rgb(hex)
    lin <- linearize(rgb)
    0.2126 * lin[1] + 0.7152 * lin[2] + 0.0722 * lin[3]
  }
  l1 <- relative_luminance(hex1)
  l2 <- relative_luminance(hex2)
  (max(l1, l2) + 0.05) / (min(l1, l2) + 0.05)
}

# Example: is a purple accent readable as small text on a light card
# background? WCAG needs >= 4.5 for normal text, >= 3 for large text
# (18pt+ regular or 14pt+ bold).
wcag_contrast("#6A4C93", "#EDE7F6")
```

Check every foreground/background pairing your new color actually
appears in, at the text size it's actually used at (a `text-box` section
label is small text and needs 4.5:1; a 28pt stat-card number is large
text and only needs 3:1). This project's own `default.typ` and `uk.typ`
theme files have real, first-hand examples of a color that passed at one
size and failed at another; both files' comments explain exactly which
pairing failed and why.

### Rendering with a custom theme

```{r, eval = FALSE}
render_onepager(data, template = "trend_snapshot", theme_path = "my-theme.typ", output = "report.pdf")
```

## Part 2: Logos and images

### Logos are data, not template edits

Every built-in template reads these tokens from the `data` list you pass
to `render_onepager()`. Only the primary logo is required:

| Token | Meaning |
|---|---|
| `logo_primary_path`, `logo_primary_alt` | Required. Your own organization's logo, always shown |
| `logo_partner_a_path`, `logo_partner_a_alt`, `show_partner_a` | Optional, off by default. A co-branding partner, shown left of the primary logo |
| `logo_partner_b_path`, `logo_partner_b_alt`, `show_partner_b` | Optional, off by default. A second co-branding partner, shown right of the primary logo |
| `header_texture_path` | Optional, defaults to onepagr's bundled texture. The decorative background behind the header band |

Changing a logo means changing these values plus staging the actual
image file, via `render_onepager()`'s `extra_assets` argument; it does
**not** require exporting or hand-editing a template's Typst source.
Three separate logo slots exist (rather than one flattened composite
image) so a screen reader announces real per-organization identification
instead of one opaque "logo", but not every jurisdiction has a
three-organization design. `show_partner_a`/`show_partner_b` are each
independent `"true"`/`"false"` toggles (the literal lowercase string, not
an R logical; see `?render_onepager`) that default to `"false"`, so a
single health department reporting under its own name passes only the
primary logo and gets one logo with no dangling divider. A two-agency
partnership switches on one partner, and a KIPRC-style three-organization
lockup switches on both. The primary logo is never optional. If you
switch a partner on, you must also supply its path and alt text:
onepagr stops with an error rather than render a blank or unlabeled logo.

```{r, eval = FALSE}
# A single health department: one logo, nothing to switch off.
data$logo_primary_path <- "county-health-dept-logo.png"
data$logo_primary_alt <- "Example County Health Department logo"

render_onepager(
  data, template = "syndromic_alert", theme = "default", output = "report.pdf",
  extra_assets = "county-health-dept-logo.png"
)

# A three-organization lockup: switch on both partners and supply each
# one's path and alt text.
data$logo_partner_a_path <- "partner-a-logo.png"
data$logo_partner_a_alt <- "Partner A logo"
data$show_partner_a <- "true"
data$logo_partner_b_path <- "partner-b-logo.png"
data$logo_partner_b_alt <- "Partner B logo"
data$show_partner_b <- "true"

render_onepager(
  data, template = "syndromic_alert", theme = "default", output = "report.pdf",
  extra_assets = c(
    "county-health-dept-logo.png", "partner-a-logo.png", "partner-b-logo.png"
  )
)
```

`extra_assets` stages your files into the same directory Typst compiles
from (Typst's compiler sandboxes file access and rejects absolute paths
outright), so the token value must be just the basename you pass to
`extra_assets`, not the file's original full path. If you're fine with
the package's own three built-in placeholder logos, you don't need
`extra_assets` at all: the fixture defaults in
`vignette("getting-started")` already point at those bundled assets by
name.

### A practical tip: crop before you resize

A raster logo file often has transparent padding baked in around the
real artwork (this project found one source logo with roughly a third
of its canvas being empty space). That padding silently shrinks how big
the logo actually looks once Typst applies a `height:`/`width:` setting,
and can misalign a multi-logo row that assumes each image's visible
content fills its frame. Check the image's real bounding box before
trusting a nominal size setting:

```{r, eval = FALSE}
# Using the PIL/Pillow Python library's alpha.getbbox(), or an
# equivalent in your tool of choice, before sizing a new logo.
```

### Accessible alt text

Every logo `image()` call takes an `alt:` argument. Write real,
specific text ("Sample Health Department logo"), not something generic
like "logo" or "image": this is what a screen reader announces in
place of the image, and it's the same accessibility bar the rest of
onepagr's output is held to.

## Part 3: Fonts

A theme's `body-font` key names the font family Typst should use for all
body/heading text (e.g. `default.typ`'s `body-font: "Liberation Sans"`).
It can also be a list of fallbacks that Typst tries in order, which is how
`kdph.typ` asks for Calibri, then Carlito (an open font with identical
metrics), then Liberation Sans: `body-font: ("Calibri", "Carlito",
"Liberation Sans")`. A family missing from the list produces a warning,
not an error, as long as a later one is found.
Two different questions hide inside "change the font":

1. **What family name should Typst look for?** That's the theme's job:
   change `body-font` in your own theme file, same as any other token in
   Part 1.
2. **Is that family actually available to Typst at compile time?** That's
   a separate, per-render concern, and it's the one that actually needs
   new machinery: Typst only sees fonts installed system-wide by default,
   and a locked-down enterprise deployment (e.g. Posit Workbench) may not
   permit installing new fonts at the OS level at all.

`render_onepager()`'s `font_dir` argument solves the second problem
without needing OS-level font installation: point it at a directory
containing your own `.ttf`/`.otf` files, and Typst searches that
directory *in addition to* system fonts for this compile only
(confirmed directly: `quarto typst fonts --font-path <dir>` lists a font
from `<dir>` alongside every system font, not instead of them).

```{r, eval = FALSE}
# my-theme.typ sets body-font: "Open Sans", and Open Sans isn't
# installed system-wide: ship the font file alongside your project
# and point font_dir at its folder.
render_onepager(
  data, template = "cohort_summary", theme_path = "my-theme.typ",
  output = "report.pdf", font_dir = "fonts/"
)
```

If `body-font` names a family Typst can't find (system-wide or in
`font_dir`), Typst silently falls back to a default font rather than
erroring, so after changing a font, actually open the compiled PDF (or
check its embedded font list) to confirm your font took effect, the same
"verify, don't assume" discipline Part 1 applies to color contrast.

## Onboarding a new jurisdiction: what you actually need

Pulling Parts 1-3 together, a health department or university adopting
onepagr for the first time needs exactly three things before it can
render its own reports, on top of the same real analysis output every
other onepagr user needs:

1. **A color palette.** At minimum, a primary/accent color pair; see
   Part 1. Reuse the built-in `default` (Bootstrap-derived, brand-neutral)
   theme as-is if you don't have (or don't yet want to commit to)
   your own institutional colors. A full custom theme is colors, not
   code: copy `default.typ`, change the hex values, verify contrast, done.
2. **Logo file(s).** One, if you're reporting under your own name alone;
   two or three if you co-brand with partner agencies. See Part 2:
   there's no minimum of three, and no template editing required either
   way.
3. **A font, if the default doesn't suit you.** Optional: every
   built-in theme already names a real, commonly-available font. Only
   relevant if your institution has a specific brand font it wants to
   use instead; see Part 3.

None of the three requires touching a template's Typst source. A new
jurisdiction's entire setup is a theme file (or none, if `default`/`uk`
suffice), a handful of logo files, and the `data` list documented in
`vignette("getting-started")` and `vignette("end-to-end-workflow")`.

## Part 4: Type size and spacing

Three theme keys control type and spacing. Each has a no-op default, so a
built-in theme renders exactly as documented until you change one.

| Key | Meaning |
|---|---|
| `min-font-size` | No text renders smaller than this. `0pt` means no floor. |
| `font-scale` | Multiplies every text size (and the boxes and label columns that hold text). `1.0` is unchanged. |
| `space-scale` | Multiplies padding, gaps, and vertical spacing. `1.0` is unchanged. |

You can also set any of them for one report without touching a theme, through
optional data tokens (`min_font_size` in points, `font_scale`, `space_scale`),
passed as strings such as "12":

```{r, eval = FALSE}
# No text smaller than 12pt, for a large-print version.
render_onepager(
  c(data, list(min_font_size = "12")),
  template = "trend_snapshot", theme = "default", output = "large-print.pdf"
)
```

A per-render value wins over the theme's. An unset one falls back to it.

The floor only raises text that is below it and leaves larger text alone.
It does not scale everything, so with a 12pt floor the smaller sizes (7 to
9pt in the built-in templates) all become 12pt and weight and color carry the
hierarchy. `font-scale` is the proportional alternative: it keeps every size
relationship and grows the layout with it. The floor governs text the
templates typeset. It cannot change text baked into a map image or a logo you
supply.

The built-in fixed-page templates are designed to fill exactly 2 pages at the
defaults, and are tested that way under every built-in theme. Raising type or
spacing can push them past 2 pages (a 10pt floor takes `cohort_summary` to 4
pages and `county_choropleth` to 3).
onepagr allows that and prints a message when it happens, for example
"cohort_summary is designed for 2 pages; this render produced 4." Extra pages
of a fixed template have no footer, because the footer is placed once on each
designed page. A render that has been scaled or floored is outside the
package's accessibility verification, so run PAC on the result.

Boxes that hold text grow with the type: label columns, the map explainer box
in `county_choropleth`, and the reserved bottom margin that holds the footer
in the alert templates. Map images, logos, and the header texture are
fixed-size and do not. After a floored or scaled render, look it over for
text that runs past its box, and run PAC on it.

## Part 5: Rewording headings, labels and text

Every piece of reader-facing text in a built-in template is an optional data
token: section headings, box labels, captions, paragraphs, bullets, bar-chart
row labels, the metadata strip and footer labels, and the alt text for charts
and maps. Each one defaults to the text the template ships with, so you supply
only the ones you want to change, and a template you leave alone renders
exactly as before. You never need to edit Typst.

`template_tokens()` lists everything a template takes. Required tokens (your
numbers and names) come first, then the optional ones with their current text:

```{r}
tokens <- onepagr::template_tokens("overdose_spike_alert")
tokens[!tokens$required, c("token", "default")]
```

Optional tokens are named by what they set, so you can filter on the prefix:

| Prefix | Sets |
|---|---|
| `heading_*` | Section headings |
| `label_*`, `banner_*` | Box, group and alert-banner labels |
| `stat_*` | Captions under big numbers |
| `bar_*` | Row labels in bar charts |
| `text_*` | Paragraphs, bullets and notes |
| `alt_*` | Alt text for charts and maps |
| `strip_label_*`, `footer_label_*` | Labels in the metadata strip and footer |
| `chips_*` | A short list of items, separated by `|` |

```{r}
cohort <- onepagr::template_tokens("cohort_summary")
head(cohort[grepl("^heading_", cohort$token), c("token", "default")], 4)
```

Use them like any other value:

```{r, eval = FALSE}
render_onepager(
  c(data, list(
    heading_alert_details = "WHAT WE KNOW",
    label_actions = "WHAT TO DO NEXT",
    text_threshold_note = "Alert level: {{{threshold}}} overdoses in 24 hours."
  )),
  template = "overdose_spike_alert", theme = "default",
  output = "reworded.pdf"
)
```

A few things to know:

* A name that no template uses is ignored with a warning that suggests the
  closest token, such as `"heading_glnce" was ignored ... Did you mean
  "heading_glance"?`. A name that belongs to a different built-in template
  is left alone, so you can share one data list across templates.
* A value is set as Typst markup. `*bold*` and `_italic_` work. To show `@`,
  `$`, `*`, `_`, `#` or a backtick literally, put a backslash before it.
  Alt text (`alt_*`) is plain text, and should not contain a double quote.
* The number in a default is filled from your data. A default such as
  `Results (N = {{{n_total}}})` uses your `n_total`. Your own replacement is
  used exactly as written, so type any numbers into it yourself. The token in
  the default is still required, because the template needs it whether or
  not you keep the default text.
* Headings stay real headings in the PDF's structure whatever text you give
  them, so screen readers and the document outline keep working. Update the
  `alt_*` text whenever you change what a chart or map shows.
* A longer heading or paragraph can wrap onto more lines and push a
  fixed-length template past its page count. Render and look after
  rewording.
* To change a template's structure rather than its words, such as adding a
  section, use `export_template()` and edit the Typst.

## See also

- `vignette("getting-started", package = "onepagr")` for the basic
  render workflow.
- `vignette("end-to-end-workflow", package = "onepagr")` for a full
  worked example starting from real data.
- Any file under `system.file("typst/themes", package = "onepagr")`
  for a complete, real theme file to read or copy.
